Values disclosures and trust in science: A replication study

### 2022-07-15 paper/paper.Qmd
99% first draft

*[Anonymized original data and reproducible code are available at <https://github.com/dhicks/transparency>.  Instructions in that repository explain how to automatically reproduce our analysis.]*

@@

Critically, our data are unlikely to be representative by education level and political ideology.  In 2021, about 9% of US adults 25 or over had a less than high school education, and 38% had a Bachelor's degree or higher [@CPSHistoricalTime2022 fig. 2].  Only 1% of our participants reported a less than high school education, and 57% reported a Bachelor's degree or higher.  For political ideology, the General Social Survey has consistently found over several decades that about 30% of US adults identify as liberal, about 30% identify as conservative, and about 40% as moderate [@GSSDataExplorer2022].  Among our participants, liberals (574) heavily outnumber conservatives (248; @fig-part-values).  Both overrepresentation of college graduates and underrepresentation of conservatives (especially conservatives with strong anti-institutional views) are known issues in public opinion polling [@KennedyEvaluation2016Election2018].  In exploratory data analysis, we noted that there was essentially no correlation between political ideology and perceived trustworthiness *[supplement]*.  This was surprising, since general trust/distrust in science has become a partisan phenomenon over the last few decades:  data from the General Social Survey shows increasing trust in science from liberals and decreasing trust from conservatives [@GauchatPoliticizationSciencePublic2012; @LeePartyPolarizationTrust2021], and many (though not all) prominent public scientific controversies align with liberal-conservative partisanship [@FunkAmericansPoliticsScience2015].

@@

*[maybe move to conclusion]*
The evidence here suggests that if a scientist discloses their values alongside their presentation of scientific research, what values they share predicts their perceived trustworthiness, and that a scientist who discloses valuing public health will tend to be perceived as more trustworthy than the scientist who discloses valuing economic growth.

@@

We did not find evidence for a transparency penalty or shared values effect.  And we found evidence of a scientist values effect, which also did not appear in @ElliottValuesEnvironmentalResearch2017.

@@

The disagreement over a transparency penalty is more difficult to explain.  Using our approach and the data from @ElliottValuesEnvironmentalResearch2017, the estimate for a values disclosure was statistically significant (@tbl-bc model 1).  One possible explanation is that, over the course of the Covid-19 pandemic, members of the general public have become used to "scientists" (including both bench researchers and public health officials) making claims about the importance of protecting public health.  (And, more recently, Leana Wen and some other prominent physicians and public health officials have argued for the importance of a "return to normal," which might be understood as prioritizing economic growth.)  A values disclosure that was regarded as violating the value-free ideal, pre-pandemic, might now be seen as routine.

*[further work]*

>>>>

Anonymized original data and reproducible code are available at <https://github.com/dhicks/transparency>.  Instructions in that repository explain how to automatically reproduce our analysis.

@@

Critically, our data are unlikely to be representative by education level and political ideology.  In 2021, about 9% of US adults 25 or over had a less than high school education, and 38% had a Bachelor's degree or higher [@CPSHistoricalTime2022 fig. 2].  Only 1% of our participants reported a less than high school education, and 57% reported a Bachelor's degree or higher.  For political ideology, the General Social Survey has consistently found over several decades that about 30% of US adults identify as liberal, about 30% identify as conservative, and about 40% as moderate [@GSSDataExplorer2022].  Among our participants, liberals (574) heavily outnumber conservatives (248; @fig-part-values).  Both overrepresentation of college graduates and underrepresentation of conservatives (especially conservatives with strong anti-institutional views) are known issues in public opinion polling [@KennedyEvaluation2016Election2018].  In exploratory data analysis, we noted that there was essentially no correlation between political ideology and perceived trustworthiness *[supplement?]*.  This was surprising, since general trust/distrust in science has become a partisan phenomenon over the last few decades:  data from the General Social Survey shows increasing trust in science from liberals and decreasing trust from conservatives [@GauchatPoliticizationSciencePublic2012; @LeePartyPolarizationTrust2021], and many (though not all) prominent public scientific controversies align with liberal-conservative partisanship [@FunkAmericansPoliticsScience2015].

@@

@@

We did not find evidence for a transparency penalty or shared values effect.  And we found evidence of a scientist values effect (a scientist who discloses valuing public health is perceived as more trustworthy), which was not identified by @ElliottValuesEnvironmentalResearch2017.

@@

The disagreement over a transparency penalty is more difficult to explain.  Using our approach and the data from @ElliottValuesEnvironmentalResearch2017, the estimate for a values disclosure was statistically significant (@tbl-bc model 1).  One possible explanation is that, over the course of the Covid-19 pandemic, members of the general public have become used to "scientists" (including both bench researchers and public health officials) making claims about the importance of protecting public health.  A values disclosure that might have been regarded as violating the value-free ideal, pre-pandemic, might now be seen as routine.

The scientist values effect — a scientist who discloses valuing public health is perceived as more trustworthy — might be interpreted as supporting some of the claims of the "aims approach" in philosophy of science.  This approach argues that scientific fields often have both social or practical aims, along with epistemic aims such as the pursuit of truth [@ElliottNonepistemicValuesMultiple2014; @IntemannDistinguishingLegitimateIllegitimate2015; @PotochnikIdealizationAimsScience2017; @FernandezPintoLegitimizingValuesRegulatory2019; @HicksWhenVirtuesAre2022].  For example, the field of public health might have the practical aim of promoting the health of the public.  Where certain values are seen as conflicting with these aims, it is morally wrong for a scientist to promote those values.  Rather than violating the value-free ideal, a scientist who discloses valuing public health might be seen as following the norms of the field of public health.

## Limitations

As noted above, a major limitation of the current study is that the sample substantially underrepresents political conservatives.  Our a priori DAG analysis and empirical robustness checks do not indicate any bias in our reported estimates due to this representation problem.  However, our estimates could still be biased if any of the effects examined here have interactions with political ideology.

Another limitation is that our sample only considers US adults.  It does not include residents of any other country, and does not track where our participants might live within the US.  Public scientific controversies often have different dynamics both across national borders and within different regions of the US.

A third limitation is that the stimulus only considers a single controversy, over the safety of the chemical bisphenol A (BPA).  Public attention to this controversy peaked around 2009.  While it remains unsettled in terms of both science and policy, it is much less socially and politically salient than controversies over climate change or vaccination.

## Directions for future work

We suggest three directions for future work in this area, two in the experimental stimulus and one in the endpoint or outcome measured.

The stimulus developed by @ElliottValuesEnvironmentalResearch2017 involves a single (fictional) scientist.  But public scientific controversies often feature conflicting claims made by contesting experts and counterexperts.  For example, as the Omicron wave of the Covid-19 pandemic waned in the US in spring 2022, various physicians, public health experts, journalists, and government officials made conflicting claims about the effective severity of Omicron, the relative effectiveness of masks in highly vaccinated populations, and the need to "return to normal" [@Adler-BellPandemicInterpreter2022; @KhullarWillCoronavirusPandemic2022; @YongHowDidThis2022].  For members of the general public, the question was not whether to trust the claims of a given expert, but instead which experts to trust.  Publics might perceive both expert A and expert B to be highly trustworthy, but would need to make a decision about who to trust if these experts are making conflicting claims.  It would be straightforward to modify the single-scientist stimulus to cover this kind of "dueling experts" scenario.  To modify METI, participants might be asked which of the two experts they would consider more competent, ethical, honest, etc.

@BrownTrustExpertiseScientific2022 emphasized a distinction between individuals, groups, and institutions, as both trustors and trustees, and argued that much of the philosophical and empirical research on trust in science has focused on either individual-individual trust (as in @ElliottValuesEnvironmentalResearch2017) or individual-group trust (as in the General Social Survey, which asks individual respondents about their confidence in "the scientific community").  In future experiments, "Dr. Riley Spence" could be represented with an institutional affiliation, such as "Dr. Riley Spence of the Environmental Protection Agency" or "Dr. Riley Spence of the American Petroleum Institute" Or claims could be attributed directly to institutions, such as "Environmental Protection Agency scientists" or "a report published by the American Petroleum Institute."

Finally, research on the climate controversy suggests that acceptance of scientific claims might underpredict policy support.  Since 2008, public opinion studies by the Yale Program on Climate Change Communication have found that only about 50-60% of the US public understands that there's a scientific consensus on climate change, that this understanding is politically polarized (conservatives are substantially less likely to recognize the consensus), but also that some climate policies enjoy majority support even among conservatives [@LeiserowitzPoliticsGlobalWarming2022].  Future research could ask participants whether they would support, for example, increased regulation of BPA, either instead of or along with the perceived trustworthiness of the scientific expert.
